Coordinating Distributed SDN Controllers\\{\Large On the Synchronization Power of OpenFlow

### 2015-03-07 inband-STN.tex
stefan

\title{On the Synchronization Power of OpenFlow}

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are typically \emph{distributed systems}: to achieve availability,
load balancing and responsiveness, control modules are installed on
multiple machines.   
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins. 
%[[PK this is a contraversial point, not important for us 
%Moreover, an SDN may be managed by multiple administrators simultaneously. 
%]]
Unfortunately, these goals are hard to combine with consistent network
manipulation and certain synchronization challenges among the
distributed controllers must be addressed. 
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane. 
%]]
In this paper, we show that powerful synchronization mechanisms
can be implemented \emph{in band}, simply using today's Openflow switches.
We show that standard OpenFlow features allow for providing simple
but general-purpose transactions with all-or-nothing semantics. 
We then show that these transactions can be used to reach consensus
among software controllers and, even more surprizingly, to resolve
\emph{on-the-fly} conflicts between concurrent policy updates in an
effectively atomic way, without affecting the ongoing traffic.  
%[[PK
%In particular, we describe the implementation of test\&set primitives 
%as well as multi-writer consensus objects which can be used
%to implement atomic transactions with all-or-nothing semantics.
%Interestingly, our approach allows controllers to detect and
%resolve very general notions of conflicts, not only, e.g., conflicts
%due to overlapping flow spaces but also conflicts based on network load.
%]]
\end{abstract}

@@

By consolidating and outsourcing the control over the dataplane switches to a logically
centralized controller, the concept Software-Defined Networking (SDN)
simplifies network management and facilitates faster innovations, 
as the (software) control plane is decoupled from the (hardware) data
plane and can evolve independently.
%[[PK do not see why it should be here
%Moreover, Openflow, the standard SDN protocol today, introduces many generalizations,
%in terms of traffic engineering, definition of flows, as well as in-band network functionalities,
%by relying on a simple match-action paradigm which allows us to define
%forwarding rules based not only on Layer-2, but also Layer-3 and Layer-4 (and possibly beyond)
%header fields.
%]]
%
The distinct network control plane has the potential to simplify
network management
%[[PK did not get
%---many network tasks are inherently non-local---
%]]
and opens new opportunities for innovation,
%[[PK we rather  talk about implementing it? 
%managing and operating an SDN in 
but implementing it 
%]]
%[[PK adaptive to what?
%a more adaptive but 
%]]
in a scalable and  consistent manner is highly non-trivial.
%[[PK did not get ``even''
%, even with SDN.
%]]

Indeed, there is a wide consensus that to achieve availability,
responsiveness, and balancing of network management tasks, 
the \emph{logically} centralized SDN
control plane must be physically \emph{distributed}.
First, in order to provide a high availability as well as a minimal degree of
fault-tolerance, controllers should be redundant~\cite{onix,stn,onos}: a failure
of one controller can be masked by other controllers. Second, it has also been proposed
to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication intensive control plane events close to their origin.~\cite{devoflow,kandoo,jukka,disco}
Third, larger SDNs are likely to be operated by multiple administrators or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory,stn}.
All this, however, comes with the need for a \emph{synchronization}
between distributed controllers, assuming that we want them to
manipulate the network \emph{consistently}~\cite{cpc}.
And this task is highly nontrivial~\cite{cap-theorem}.

@@

Today, unfortunately, we do not have a good understanding yet of how to realize
such distributed control planes. The problem is essentially a
distributed-systems 
one: Multiple controllers may simultaneously try to
install conflicting updates and we want to resolve these conflicts
\emph{consistently} (no undesired behavior is observed on the data
plane) and \emph{efficiently} (no undesired delays are imposed on the
control application).

Consider, for example, the particularly interesting problem of
consistent installation of new forwarding policies, stipulating routes
that packets of different header spaces should follow across the
network~\cite{network-update,roger-hotnets,correct,stn}.
Installing conflicting forwarding rules, e.g., rules of the same priority defined over non-disjoint
flow spaces may lead to pathological network behavior (loops,
bleckholes, etc.).
Similarly, installing diverging load-balancing policies may, when combined, \emph{increase} the load~\cite{log-cent}. 
To render things more difficult, controllers may also fail, even before their updates have been completed.

%\subsection{In band synchronization?}
%
In this paper, we look closely at the most popular specification
language for SDN control-data plane exchange -
OpenFlow~\cite{of-spec}. 
At a high level, the specification represents the exchange in the form
of \emph{control messages}, containing commands for installing or
modifying \emph{rules} that, in turn, constitute the \emph{flow table} of
a data-plane switch used in processing the ongoing data-plane traffic.

We observe that the standard provides the option of \emph{bundling}
these messages in one, so that either all of the control messages are
processed or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled messages cannot
be processed, all the other requests  are rejected and an error
message is sent to the controller that issued them.       
Interestingly, this bundled processing can be, when properly configured, done in a
\emph{packet-atomic} way, so that each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
request was processed.

Intuitively, if we treat the flow table at the switch as a \emph{shared memory}
that can be concurrently read and modified by multiple controllers, 
the bundling features can be seen as a \emph{multi-write} or, more
widely, \emph{multi-modify} operation that modifies
multiple memory location in one atomic step.
This ability is reminiscent of the hypothetical multi-write
shared-memory model which is known to enable solutions for important
synchronization problems, such as consensus, impossible to solve
otherwise~\cite{multi-objects}. 
Indeed, we describe an algorithm that solves consensus among any
number of controllers using a single OpenFlow switch that accepts
bundle control messages.

Even more, we observe that the standard also assumes \emph{conditional} application of
certain types of control messages. A specific $\textsf{OFPFF\_CHECK\_OVERLAP}$ flag
set in the control message containing a new forwarding rule ensures
that the rule is applied if and only if it does not \emph{conflict}
with any other rule currently in the flow table.  
This in-band conflict detection mechanism, combined with bundling,
allows us to implement specific types of \emph{transactions} that may
atomically install multiple rules if certain compatibility conditions
are met.

We then discuss how these synchronization mechanisms can be used to
achieve consistent network management. 
We first show that they can be used to replace the read-modify-write
primitives voluntarily assumed in the recent implementation of a
recent proposal for a consistent composition system for forwarding policies~\cite{cpc}.  
We then use these transactions directly to implement a lightweight
forwarding policy composition system that requires no additional
synchronization on the control plane.     
We then explore how more general classes of policies can be composed
\emph{on-the-switch}.

Our results highlight the surprising synchronization power hidden in the
OpenFlow standard. The synchronization mechanisms we describe in this
paper turn out to be crucial for ensuring consistency of network
management on the control plane while preserving consistency of the
traffic processed on the data plane.

@@

%\paragraph{Organization.}
%
The rest of the paper is organized as follows.
In Section~\ref{sec:model}, we overview the basics of the  OpenFlow
protocol and recall the bundle and \textsf{ofpff\_check\_overlap} features. 
In Section~\ref{sec:sync}, we specify the implementations of our OpenFlow-based
sycnhronization mechanisms. 
In Section~\ref{sec:application}, we discuss a few policy composition
algorithms that benefit from these mechanisms. 
In Section~\ref{sec:conclusion}, we discuss open questions and related work.

@@

\section{SDN primer and OpenFlow features}\label{sec:model}

Define the protocol (from Reclaiming the Brain"?) and discuss bundling
and check-overlap.

@@

\subsection{Improving the scheme with check-overlap}\label{sec:todo}
Till now we considered a model that allows multiple updates to match-action entries. Here we extends the model with actions create-entry and delete-entry that may failed and abort the whole transaction depending on the existence on non-existence of specific entries prior to the action.
%TBD - much less resources. easily solving the infinite versions issue.

>>>>

%\title{On the Synchronization Power of OpenFlow}

\title{Coordinating Distributed SDN Controllers\\{\Large On the Synchronization Power of OpenFlow}}

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads, 
control modules are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, a distributed control plane also introduces
new challenges, and in order to guarantee a consistent network operation,
the actions of different controllers may need to be synchronized and coordinated.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
In this paper, we show that powerful synchronization mechanisms
can be implemented \emph{in band}, simply using today's Openflow switches.
We show how to leverage standard OpenFlow features for providing simple
but general-purpose transactions with all-or-nothing semantics.
We then show that these transactions can be used to reach consensus
among software controllers and, what is more, to resolve
general conflicts between concurrent policy updates in an
effectively atomic way and \emph{on-the-fly}, without affecting the ongoing traffic.
%[[PK
%In particular, we describe the implementation of test\&set primitives
%as well as multi-writer consensus objects which can be used
%to implement atomic transactions with all-or-nothing semantics.
%Interestingly, our approach allows controllers to detect and
%resolve very general notions of conflicts, not only, e.g., conflicts
%due to overlapping flow spaces but also conflicts based on network load.
%]]
\end{abstract}

@@

By consolidating and outsourcing the control over the dataplane switches to a logically
centralized controller, the concept Software-Defined Networking (SDN)
simplifies network management and facilitates faster innovations,
as the (software) control plane is decoupled from the (hardware) data
plane and can evolve independently.
Moreover, Openflow, the standard SDN protocol today, introduces many generalizations,
in terms of traffic engineering, definition of flows, as well as in-band network functionalities,
by relying on a simple match-action paradigm which allows us to define
forwarding rules based not only on Layer-2, but also Layer-3 and Layer-4 (and possibly beyond)

While the logically centralized perspective offered by SDN is attractive,
there is a wide consensus that
the control plane should be physically \emph{distributed}. 
First, in order to provide a high availability as well as a minimal degree of
fault-tolerance, controllers should be redundant~\cite{onix,stn,onos}: a failure
of one controller can be masked by other controllers. Second, it has also been proposed
to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication intensive control plane events close to their origin.~\cite{devoflow,kandoo,jukka,disco}
Third, larger SDNs are likely to be operated by multiple administrators or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory,stn}.
All this, however, comes with the need for \emph{synchronization}
and \emph{coordination} among the distributed controllers:
the network should be manipulated \emph{consistently}~\cite{cpc}.
This task is nontrivial~\cite{cap-theorem}.

@@

Today, we do not have a good understanding yet of how to realize
such distributed control planes. The problem is essentially a
distributed systems
one: Multiple controllers may simultaneously try to
install conflicting updates and we want to resolve these conflicts
\emph{consistently} (no undesired behavior is observed on the data
plane) and \emph{efficiently} (no undesired delays are imposed on the
control application).

Consider, for example, the practical problem of
consistent installation of new forwarding policies, stipulating routes
that packets of different header spaces should follow across the
network~\cite{network-update,roger-hotnets,correct,stn}.
Installing conflicting forwarding rules, e.g., rules of the same priority defined over non-disjoint
flow spaces may lead to pathological network behavior (loops,
blackholes, etc.).
Similarly, installing diverging load-balancing policies may, 
when combined, \emph{increase} the load~\cite{log-cent}.
To render things more difficult, controllers may also fail, 
even before their updates have been completed.

%\textbf{} 
\paragraph{Our Contributions}
In this paper, we show that 
OpenFlow~\cite{of-spec}---the standard SDN protocol today---can
be used to provide strong controller synchronization and coordination 
primitives \emph{in-band}.
In particular, we show how to exploit the \emph{bundling}
feature and the $\textsf{OFPFF\_CHECK\_OVERLAP}$ flag,
to implement a \emph{shared memory} abstraction of the switch
and its flow tables: 
This shared memory 
allows us to implement
\emph{transactions} which may
atomically install multiple rules if certain compatibility conditions
are met, and which also facilitate \emph{packet-atomic} rule updates. 
These transactions can be \emph{multi-write} or even \emph{multi-modify}:
while not supporting
read operations, the transaction can be conditioned on the presence of
certain rules.
This ability is reminiscent of the hypothetical multi-write
shared-memory model which is known to enable solutions for important
synchronization problems, such as consensus, impossible to solve
otherwise~\cite{multi-objects}.
Indeed, we describe an algorithm that solves consensus among any
number of controllers using a single OpenFlow switch that accepts
bundle control messages.

Our results highlight the synchronization power hidden in the
OpenFlow standard. The synchronization mechanisms we describe in this
paper turn out to be crucial for ensuring consistency of network
management on the control plane while preserving consistency of the
traffic processed on the data plane.
As a case study, we also show how to replace the read-modify-write
primitives voluntarily assumed in the recent implementation of a
consistent composition system for forwarding policies~\cite{cpc}.
We also show how to use these transactions directly to implement a lightweight
forwarding policy composition system that requires no additional
synchronization on the control plane.
Finally, we explore how more general classes of policies can be composed
\emph{on-the-switch}.

@@

\paragraph{Organization}
The rest of the paper is organized as follows.
In Section~\ref{sec:model}, we overview the basics of the  OpenFlow
protocol and recall the bundle and \textsf{ofpff\_check\_overlap} features.
In Section~\ref{sec:sync}, we specify the implementations of our OpenFlow-based
sycnhronization mechanisms.
In Section~\ref{sec:application}, we discuss a few policy composition
algorithms that benefit from these mechanisms.
In Section~\ref{sec:conclusion}, we discuss open questions and related work.

@@

\section{OpenFlow Background}\label{sec:model}

The synchronization and coordination mechanisms presented in this paper
are based on Openflow~\cite{of-spec}, the standard SDN protocol today. In this section, 
we provide the necessary background.

es a small partial deployment of SDN and
out-sources the control over a set of Openflow switches to a logically centralized software controller.
Openflow, the standard SDN protocol today, is based on a match-action paradigm: the controller
can install simple rules on the Openflow switches.

In a nutshell, 
OpenFlow
is based on a match-action concept: OpenFlow switches store
rules (installed by the controller) consisting of a match and an
action part. A packet matched by a certain rule will be subject
to the associated action (to be immediately applied now or later). 
For example, an action can define a port to which the
matched packet should be forwarded, or add or change a tag
(a certain part in the packet header).
Each OpenFlow switch stores one or multiple flow tables,
each of which contains a set of rules (flow entries). Flow
tables form a pipeline, and flow entries are ordered according
to priorities: A packet arriving at a switch is first checked by
the rule of the highest priority in table 0: the fields of the
data packet are compared with the match fields of that rule,
and if they fit, some instructions (the actions) are executed;
subsequently, lower priority rules are checked. Depending on
the outcome of the table 0 processing, the packet may be sent
to additional flow tables in the pipeline; during the traversal of
the flow tables, the packet can be equipped with meta-tags to
transfer temporary information. Concretely, instructions can be
used to define additional tables to be visited (goto instruction),
to modify the set of to-be-applied actions (either by appending,
deleting, or modifying actions), or immediately apply some
actions to the packet. A meta-data field can be used to exchange
information between tables; it is part of the header that can be
inserted or removed from a packet via push and pop actions.
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
(For instance, the meta-data field is maskable.) If no rule
matches, the packet is dropped.

\textbf{TODO LIRON: what else is missing? describe cookies?}

The Openflow standard (as of \textbf{TODO}) offers two interesting
features which we will exploit to implement the transactional memory abstraction:

\begin{enumerate}
\item \textbf{Bundling:}
Control messages are aggregated into one message, so that either all of the control messages are
processed, or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled messages cannot
be processed, all the other requests  are rejected and an error
message is sent to the controller that issued them.
Interestingly, this bundled processing can be, when properly configured, done in a
\emph{packet-atomic} way, so that each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
request was processed.

\item \textbf{Conditional Application:}
The Openflow standard also supports the \emph{conditional} application of
certain types of control messages. A specific $\textsf{OFPFF\_CHECK\_OVERLAP}$ flag
set in the control message containing a new forwarding rule ensures
that the rule is applied if and only if it does not \emph{conflict}
with any other rule currently in the flow table.
This in-band conflict detection mechanism, combined with bundling,
allows us to implement specific types of \emph{transactions} that may
atomically install multiple rules if certain compatibility conditions
are met.
\end{enumerate}

\section{Vision and Problem Statement}\label{sec:vision}

This paper shows how to implement a transactional memory abstraction using
standard Openflow switches. The shared memory abstraction of the 
flow table at the switch allows multiple controllers to atomically and 
concurrently read and modify state,
using the bundling feature to implement \emph{multi-write} or, more
widely, \emph{multi-modify} transactions. .
This ability allows us to solve consensus among any
number of controllers, using a single OpenFlow switch that accepts
bundle control messages.

These synchronization mechanisms can be used to
achieve consistent network management.
We first show that they can be used to replace the read-modify-write
primitives voluntarily assumed in the recent implementation of a
recent proposal for a consistent composition system for forwarding policies~\cite{cpc}.
We then use these transactions directly to implement a lightweight
forwarding policy composition system that requires no additional
synchronization on the control plane.
We then explore how more general classes of policies can be composed
\emph{on-the-switch}.

TODO: write vision and contribution in full detail, as an overview

@@

\subsection{Improving the scheme with check-overlap}\label{sec:todo}

Till now we considered a model that allows multiple updates to match-action entries. Here we extends the model with actions create-entry and delete-entry that may failed and abort the whole transaction depending on the existence on non-existence of specific entries prior to the action.
%TBD - much less resources. easily solving the infinite versions issue.
